---
title: "Bahamian STEM Survey"
format: html
editor: visual
execute: 
  echo: false
  warning: false
---


```{r echo= FALSE, warning= FALSE, message= FALSE}

# load required libraries
library(here)
library(readxl)
library(tidyverse)
library(fuzzyjoin)
library(flextable)

# Declare file location with `here`
here::i_am("Bahamian_STEM_Survey.qmd")


# Load survey data
survey_df <- read_excel("Exploring the STEM Pathways of Bahamians (Responses).xlsx")

# Save column names for labels 
questions <- colnames(survey_df)

# Clean up df
survey_df_clean <-  rename_with(survey_df, ~ make.names(.x)) |> 
    mutate(`Where.do.you.currently.reside.` = ifelse(grepl(`Where.do.you.currently.reside.`,
                                                         pattern = "Bahamas"),
                                                   "The Bahamas",
                                                   `Where.do.you.currently.reside.`)) |> 
   rename(
    island_high_school = On.which.island.do.did.you.attend.high.school.,
    type_high_school = What.type.of.high.school.do.did.you.attend.,
    identity = Which.of.the.following.best.describes.you.,
    current_residence = Where.do.you.currently.reside.,
    undergrad_field = What.field.is.your.undergraduate.degree.in.,
    non_stem_path = How.did.you.transition.into.a.STEM.career.without.an.undergraduate.STEM.degree.,
    education_impact = Do.you.feel.that.your.educational.background.has.impacted.your.career.progression.in.STEM.,
    stem_interest_factors_highschool = Which.of.the.following.have.contributed.most.to.your.interest.in.pursuing.a.STEM.related.degree.in.college....9,
    stem_interest_fields = Which.STEM.or.STEM.related.fields.interest.you.the.most.,
    perceived_barriers_college = What.barriers..if.any..do.you.think.will.affect.your.ability.to.pursue.a.STEM.related.degree.in.college.,
    want_stem_mentor = Would.you.be.interested.in.being.matched.with.a.mentor.in.a.STEM.field.,
    stem_interest_factors_undergrad = Which.of.the.following.have.contributed.most.to.your.interest.in.pursuing.a.STEM.related.degree.in.college....13,
    first_gen_college = Are.you.the.first.in.your.immediate.family.to.attend.college.,
    current_major = What.is.your.current.major.,
    undergrad_year = What.year.are.you.in.your.undergraduate.program.,
    degree_completion_barriers = What.barriers..if.any..are.affecting.your.ability.to.complete.your.degree.,
    had_internship = Have.you.had.an.internship.or.co.op.experience.in.a.STEM.field.,
    internship_obstacle = What.has.been.the.biggest.obstacle.to.securing.an.internship.or.co.op.,
    desired_field = What.is.your.desired.field.after.graduation.,
    desired_title = What.is.your.desired.job.title.after.graduation.,
    desired_salary = What.is.your.desired.starting.salary.range..USD..,
    have_mentor = Do.you.currently.have.a.mentor.in.your.field.,
    want_mentor = Would.you.be.interested.in.being.matched.with.a.mentor.,
    mentorship_email = If.you.would.like.to.be.contacted.regarding.potential.mentorship.opportunities..please.leave.your.email.below.,
    years_since_undergrad = How.long.ago.did.you.graduate.from.your.undergraduate.STEM.program.,
    employment_status = What.is.your.current.employment.status.,
    stem_entrepreneur = Are.you.a.STEM.entrepreneur.,
    employment_satisfaction = Are.you.generally.satisfied.with.your.current.employment.situation.,
    undergrad_major = What.was.your.undergraduate.major.,
    use_stem_skills = To.what.extent.do.you.use.the.skills.and.knowledge.from.your.STEM.degree.in.your.current.job.,
    current_salary = What.is.your.current.salary.income.range..USD..,
    bahamas_challenges = What.challenges.are.there.to.working.in.the.Bahamas.as.a.STEM.professional.,
    bahamas_attractions = What.do.you.find.most.attractive.about.working.in.The.Bahamas.,
    employer_location = Where.is.your.primary.employer.located.,
    improve_bahamas_working = What.factors.would.make.working.in.The.Bahamas.more.attractive.compared.to.your.current.employment.situation.,
    has_graduate_degree = Did.you.pursue.or.are.you.currently.pursuing.a.graduate.degree.,
    type_graduate_degree = What.type.of.graduate.degree.are.you.pursuing.have.completed.,
    can_support_students = Are.you.in.a.position.to.hire..mentor..or.offer.internships.to.Bahamian.college.students.,
    mentorship_employment_email = If.you.would.like.to.be.contacted.regarding.potential.mentorship.or.employment.opportunities..please.leave.your.email.below.,
    org_support_suggestions = Do.you.have.any.suggestions.for.how.our.organizations.can.better.support.Bahamian.students.and.professionals.in.STEM.fields.,
    timestamp = Timestamp
  ) |> 
  
  # Remove those that don't fit the survey target demo
  filter(identity != "None")
  

```
# Demographic Summary

```{r demographic-summary}
# Total participants
total_participants <- nrow(survey_df_clean)

# Participants in each demographic
survey_demo <- survey_df_clean |> 
  group_by(identity) |> 
  count()

```

There were `r total_participants` participants. 

```{r}
flextable(survey_demo)
```

# 1. Where are current Bahamian college students in STEM and STEM adjacent fields?

## Majors

The most popular majors in current college students are engineering, biological sciences, and physical sciences (which were primarily composed of student pursuing studies in marine sciences).

```{r}

# Grab current undergrads
current_undergrads_df <- survey_df_clean |>
  filter(identity == "Current undergraduate student in a STEM field")

# Counts for majors
majors_df <- current_undergrads_df |>
  rename("Major_detailed" = current_major)  |>  
  mutate(Major_detailed = ifelse(grepl(Major_detailed , 
                                       pattern = "Information Technology"),
                                    "Information Technology",
                                    Major_detailed)) |>
  distinct(Major_detailed)

# Load STEM majors from UCLA Higher ed research institute list for fuzzy matching
majors_categories <- read.csv(file = here("majors_HERI.csv"))

# Fuzzy matching for categories
# Note - if both majors listed, only one will be used
major_matches  <- stringdist_join(majors_df, majors_categories,
                by='Major_detailed', #match based on team
                mode='left', #use left join
                method = "jw", #use jw distance metric
                max_dist=99,
                distance_col='dist') |>
  group_by(Major_detailed.x) |>
  slice_min(order_by=dist, n=1) |>
  ungroup() |>
  group_by(Major) |>
  count()

# Barplot for majors 
ggplot(major_matches, aes(x = Major, y = n)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

```

## Location

Most current undergraduate students are in the Bahamas. 
Note: survey design flaw here. Did not ask where students were attending uni Some might've interpreted this as where they study. 

```{r}

# Note: survey design flaw here. Did not ask where attending uni. Some might've interpreted this differently 
location_df <- current_undergrads_df |>
    group_by(current_residence) |>
  count()

# Barplot location
# Barplot for majors 
ggplot(location_df, aes(x = current_residence, y = n)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))


```

## Years

Most students are taking the normal amount of time to complete their degrees. 

```{r}

# Get years counts
years_df <- current_undergrads_df |>
    group_by(undergrad_year) |>
  count()

# Barplot undergrad year
ggplot(years_df, aes(x = undergrad_year, y = n)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

```

# 2. Did participants have any interaction with a pre-college or summer program that piqued their interest in a STEM pr STEM adjacent field?

```{r}

highschool_and_undergrads_df <- survey_df_clean |>
  filter(identity == "Current high school student considering a STEM major"|
          identity ==  "Current undergraduate student in a STEM field")


interests_df <- highschool_and_undergrads_df |>

  # select columns related to interest in STEM
  select(identity,
         stem_interest_factors_highschool,
         stem_interest_factors_undergrad) |>

  # Bring together high school and undergrad interest columns
  mutate(contributed_interest = coalesce(stem_interest_factors_highschool,
                                         stem_interest_factors_undergrad)) |> 
  select(identity, 
         contributed_interest)

# List possible reasons of interest 
reasons_interest <- c("I attended a STEM pre-college or summer program", "High salary expectations", "I know family or community members with careers in STEM", "STEM subjects are my favorite classes in school", "Encouraged by a peer, teacher, parent or community member", "I enjoy STEM topics as a hobby/outside of the classroom", "I am good in STEM subjects", "The career I am interested in happens to be STEM")

reason_counts <- sapply(reasons_interest, function(reason) {
  sum(str_detect(interests_df$contributed_interest, fixed(interests_df$contributed_interest)))
})

interests_df_counted <- interests_df |> 
  mutate(program = as.integer(str_detect(contributed_interest, "I attended a STEM pre-college or summer program")), 
         high_salary_expectations = as.integer(str_detect(contributed_interest, "High salary expectations")),
         family_stem = as.integer(str_detect(contributed_interest, "I know family or community members with careers in STEM")),
         favorite_class = as.integer(str_detect(contributed_interest, "STEM subjects are my favorite classes in school")),
         stem_hobby = as.integer(str_detect(contributed_interest, "I enjoy STEM topics as a hobby/outside of the classroom")), 
         stem_classroom = as.integer(str_detect(contributed_interest, "I am good in STEM subjects")),
         career = as.integer(str_detect(contributed_interest, "The career I am interested in happens to be STEM")),
         encourgaed = as.integer(str_detect(contributed_interest, "Encouraged by a peer, teacher, parent or community member"))) |> 
  select(-c(contributed_interest)) |> 
  pivot_longer(!identity, names_to = "reason_interest", values_to = "number") |> 
  filter(number != 0) |> 
  group_by(identity, reason_interest) |> 
  count()

# Plot
ggplot(interests_df_counted, aes(fill=reason_interest, y=n, x=identity)) + 
  geom_bar(position='dodge', stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
        
      

```

# 3. Where are Bahamian STEM professionals? What are the factors that contribute to their location?

## Location 
Graduates of STEM degrees are mostly in the US. Some in Canada
```{r stem-professionals}
stem_professionals_df <- survey_df_clean |> 
  filter(identity == "Graduate of an undergraduate STEM program") 


abroad_df <- stem_professionals_df |> 
  count(current_residence, employment_status) |> 
  filter(employment_status == "Employed Abroad (STEM field)")

ggplot(abroad_df, aes( x =current_residence, y = n)) +
  geom_bar(stat='identity') +
  labs(x = "Locations", y = "Count")
```

## Factors contributing or working against working in the Bahamas

The primary draw to working in the Bahamas is closeness to family and cultural reasons. The primary detriment is low salary. 

```{r}

challenges_bahamas_options <- c(
  "Low Salary",
  "Job Availability",
  "Career Growth",
  "Cost of Living",
  "Quality of Life",
  "Political Stability",
  "Family Commitments",
  "Access to Healthcare",
  "Job Diversity",
  "Limited access to cutting-edge research or technology"
)

attractions_bahamas_options <- c(
  "Competitive salary and financial incentives",
  "Work-life balance and flexible work arrangements",
  "Family and community connections",
  "Lower cost of living compared to other regions",
  "Cultural and social environment",
  "Stronger career advancement and leadership opportunities",
  "Access to multiple revenue streams and entrepreneurial opportunities",
  "Increased investment in STEM research and innovation",
  "Political and economic stability",
  "Better access to housing and property ownership",
  "Nothing - I am actively looking to move abroad"
)

Bahamas_res_df <- stem_professionals_df |> 
  filter(current_residence == "The Bahamas" & !is.na(bahamas_challenges) & !is.na(bahamas_attractions)) |> 
  select(bahamas_challenges, bahamas_attractions) 
         
# Count how many responses mention each reason
challenges_reason_counts <- sapply(challenges_bahamas_options, function(reason) {
  as.integer(str_detect(Bahamas_res_df$bahamas_challenges, pattern = reason))
})

attractions_reason_counts <- sapply(attractions_bahamas_options, function(reason) {
  as.integer(str_detect(Bahamas_res_df$bahamas_attractions, pattern = reason))
})

# Count instances of challenges
bahamas_challenges_df <- cbind(Bahamas_res_df |> 
                              select(bahamas_challenges),
                            challenges_reason_counts) |> 
  pivot_longer(!bahamas_challenges, names_to = "challenge", values_to = "number") |> 
  group_by(challenge) |> 
  summarise(n = sum(number))

# Plot challenges 
ggplot(bahamas_challenges_df, aes( x =challenge, y = n)) +
  geom_bar(stat='identity') +
  labs(title = "Challenges to Working in the Bahamas", x = "Challenge", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Count instances of attractions
bahamas_attractions_df <- cbind(Bahamas_res_df |> 
                              select(bahamas_attractions),
                            attractions_reason_counts) |> 
  pivot_longer(!bahamas_attractions, names_to = "attractions", values_to = "number") |> 
  group_by(attractions) |> 
  summarise(n = sum(number))

# Plot challenges 
ggplot(bahamas_attractions_df, aes( x =attractions, y = n)) +
  geom_bar(stat='identity') +
  labs(title = "Benefits to Working in the Bahamas", x = "Benefit", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

```

## Salaries 

Many STEM professionals are in the Bahamas and being paid at the lowest end of the salary range (below 50K). Only those that work in the US reach the higher end (200K+). 

The median salaries of US and Canada were $50,909 and $45,714. The median salary of the Bahamas was not able to be estimated as the majority of participants responded that they made less tha 50K/ year.  


```{r}


# Desired salary
stem_professionals_salary <- stem_professionals_df |> 
  filter(!is.na(current_salary)) |> 
  select(current_residence, current_salary) |> 
  group_by(current_residence, current_salary) |> 
  count() |> 
  mutate(current_salary = case_when(
    current_salary %in% c("No current income",
                        "Below $50,000") ~ "Below $50,000",
    TRUE ~ current_salary
  )) |> 
  mutate(current_salary = factor(current_salary, 
                                 levels = c("Below $50,000",
                                            "$50,000 - $69,999",
                                            "$70,000 - $99,999",
                                            "$100,000 - $149,999",
                                            "$150,000 - $199,999",
                                            "$200,000 - $249,999",
                                            "$250,000 - $299,999"))) 

# Salaries of STEM graduates
ggplot(stem_professionals_salary, aes(fill =current_salary, y = n, x= current_residence)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = "Salaries by Location", x = "Salary", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Calculating median salaries by location 
bin_widths <- data.frame(bin_width = c(NA, 20000, 30000, 50000, 50000, 50000,50000), 
                         current_salary = c("Below $50,000",
                                            "$50,000 - $69,999",
                                            "$70,000 - $99,999",
                                            "$100,000 - $149,999",
                                            "$150,000 - $199,999",
                                            "$200,000 - $249,999",
                                            "$250,000 - $299,999"))

# Canada median 
canada_df <- bin_widths |> 
  left_join(stem_professionals_salary |> filter(current_residence == "Canada"), 
  by = "current_salary", keep = TRUE)
  
half_pop_can <- sum(canada_df$n, na.rm = TRUE)/2

median_group_ll_can <- 50000

cumsum_uptomedian_can <- 8 

freq_mediangroup_can <- 7 

width_mediangroup_can <- 20000

median_can <- median_group_ll_can + ((half_pop_can - cumsum_uptomedian_can ) / freq_mediangroup_can) * width_mediangroup_can  


# US median 
us_df <- bin_widths |> 
  left_join(stem_professionals_salary |> filter(current_residence == "United States of America"), 
  by = "current_salary", keep = TRUE)
  
half_pop_us <- sum(us_df$n, na.rm = TRUE)/2

median_group_ll_us <- 70000

cumsum_uptomedian_us <- 25 

freq_mediangroup_us <- 11 

width_mediangroup_us <- 30000

median_us <- median_group_ll_us + ((half_pop_us - cumsum_uptomedian_us ) / freq_mediangroup_us) * width_mediangroup_us  

# Bahamas median 
bah_df <- bin_widths |> 
  left_join(stem_professionals_salary |> filter(current_residence == "The Bahamas"), 
  by = "current_salary", keep = TRUE)
  
half_pop_bah <- sum(bah_df$n, na.rm = TRUE)/2

```

# 4. Where do Bahamians that study or have studied in STEM fields come from in the Bahamas?

Most participants that study or have studied (not high school students) come from New Providence (62%). New Providence was 73% of the population in 2022, so this number is not dissimilar to the demographics of the Bahamas (NP is not overrepresented). 

More participants from public schools come from the out islands. 

Most participants interested in STEM come from private schools.

```{r}
# Grab high school students
high_school_origin <- survey_df_clean |>
  filter(identity != "Current high school student considering a STEM major") |> 
  filter(!is.na(type_high_school) & !is.na(island_high_school)) |> 
  select(type_high_school, island_high_school) |> 
  group_by(type_high_school, island_high_school) |> 
  count()

# Plot
ggplot(high_school_origin, aes(fill=island_high_school, y=n, x=type_high_school)) + 
  geom_bar(position='dodge', stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
        

```


# 5. Do Bahamians that have studied in STEM fields end up working in their fields?

Most who studied are able to use their degrees. Lower proportion in the Bahamas than the US or Canada. Excluded our single UK participant here for brevity. 
```{r}
# People who have graduated from STEM degree
use_stem_skills_df <- survey_df_clean |>
  filter(identity == "Graduate of an undergraduate STEM program" &
           current_residence != "United Kingdom") |> 
  select(use_stem_skills, current_residence) |> 
  group_by(use_stem_skills, current_residence) |> 
  count()

ggplot(use_stem_skills_df, aes(fill=use_stem_skills, y=n, x=current_residence)) + 
  geom_bar(position='dodge', stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

```

# 6. Do salary expectations for Bahamians that are currently studying  in STEM fields meet reality in their desired work location? 

Health professions and egineering had highest expectations for salary. How do these compare to reality? Compare to salaries of professionals in question 3.

```{r}
# People who have currently studying STEM degree
salary_expectations_df <- survey_df_clean |>
  filter(identity == "Current undergraduate student in a STEM field") |> 
  select(desired_field, desired_salary) |> 
  rename(Major_detailed = desired_field)

salary_expectations_matched <- stringdist_join(salary_expectations_df, majors_categories, 
                by='Major_detailed', #match based on team
                mode='left', #use left join
                method = "jw", #use jw distance metric
                max_dist=99,
                distance_col='dist') |> 
    group_by(Major_detailed.x) |>
  slice_min(order_by=dist, n=1) |> 
  ungroup() |> 
  select(Major, desired_salary) |> 
  group_by(Major, desired_salary) |> 
  filter(!is.na(desired_salary)) |> 
  count() |> 
  mutate(desired_salary = factor(desired_salary,
                                 levels = c("Below $30,000",
                                            "$30,000 - $49,999",
                                            "$50,000 - $69,999",
                                            "$70,000 - $99,999",
                                            "$100,000+")))
  

ggplot(salary_expectations_matched, aes(fill=desired_salary, y=n, x=Major)) + 
  geom_bar(position='dodge', stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))       
```
7. How do Undergrad's perceptions of their degree change over time? (Challenges to completion)

Financial difficulties more common in first year. 


```{r}

completion_barriers_options <- c("Financial difficulties", "Academic challenges", "Lack of mentorship", "Family responsibilities", "No major obstacles")

undergrad_perceptions <- survey_df_clean |> 
  filter(identity == "Current undergraduate student in a STEM field" & !is.na(degree_completion_barriers))|> 
  select(undergrad_year, degree_completion_barriers )

# Count how many responses mention each reason
completion_barriers_count <- sapply(completion_barriers_options, function(reason) {
  as.integer(str_detect(undergrad_perceptions$degree_completion_barriers, pattern = reason))
})

undergrad_perceptions_counted <- cbind(undergrad_perceptions, completion_barriers_count) |> 
  pivot_longer(!c(degree_completion_barriers, undergrad_year), names_to = "barrier", values_to = "number") |> 
  group_by(barrier, undergrad_year) |> 
  summarise(n = sum(number))

ggplot(undergrad_perceptions_counted, aes(fill=barrier, y=n, x=undergrad_year)) + 
  geom_bar(position='dodge', stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1))  
```
